---
title: "benthic species-habitat dashboards"
author: "Tom Webb"
date: "16/03/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Producing habitat summaries and visual dashboards for European marine benthic species

The processed EMODnet benthic numerical abundance product files, now matched to sediment properties and the EMODnet broadscale seabed habitat map, are here read in and used to feed functions to provide a summary of the habitat affinities of individual named benthic species. Where habitat affinity data is available for that species from Biotic, this is included too.

First, load required packages:
```{r, message = FALSE}
library(tidyverse)
library(ggridges)
library(worrms)
library(ggtext)
library(patchwork)
```

Now, load the datasets (column types are fully specified for benth_events to avoid parsing errors):
```{r, message = FALSE}
benth_events <- read_csv(here::here(
            "data", "derived_data/benthic_abundance_sampling_events_seabed_habs.csv"),
            col_types = cols(
              datasetid = col_double(),
              datecollected = col_datetime(format = ""),
              decimallongitude = col_double(),
              decimallatitude = col_double(),
              minimumdepthinmeters = col_double(),
              sampid = col_double(),
              eventNummer = col_double(),
              MudPercent = col_double(),
              SandPercent = col_double(),
              GravelPercent = col_double(),
              TotalD50 = col_double(),
              SandD50 = col_double(),
              GravelD50 = col_double(),
              Rock10cm = col_double(),
              Rock50cm = col_double(),
              log_D50 = col_double(),
              Biozone = col_character(),
              Energy = col_character(),
              Substrate = col_character(),
              Salinity = col_character(),
              Oxygen = col_character(),
              EUNIScomb = col_character(),
              EUNIScombD = col_character(),
              Allcomb = col_character(),
              AllcombD = col_character(),
              SalcombD = col_character(),
              MSFD_BBHT = col_character()
            ))
benth_abundances <- read_csv(here::here(
  "data", "derived_data/benthic_abundances_long.csv"))
benth_taxa <- read_csv(here::here(
  "data", "derived_data/benthic_taxa.csv"))
benth_substrate_prefs <- read_csv(here::here(
  "data", "derived_data/benthic_species_substratum_prefs.csv"))
benth_substrate_key <- read_csv(here::here(
  "data", "derived_data/substrate_values_key.csv"))
```


## Getting habitat summaries for a single species

This code loads a function which takes a species Aphia ID and returns basic summary information about the benthic habitats that it has been recorded in (from the EMODnet numerical abundance product). For continuous variables (e.g. 'Percent Mud') the function returns the arithmetic mean value of the variable from all sampling events the species was recorded in, weighted by its abundance in each event. For categorical variables (e.g. 'Substrate') the function returns the relative frequency of occurrence in each catgory (again weighted by abundance). Finally, the function checks if the species is one of those occurring in our Biotic substrate dataset, and adds relevant information about its recorded habitat preferences if so.

```{r}
source(here::here("scripts", "get_species_habitats.R"))
```

An example of running for one species:
```{r}
get_species_habitats(103228)
```

## Getting habitat summaries for all
This simply runs the above function over all species:
```{r, message = FALSE}
sp_habitat_summaries <- benth_taxa %>%
  mutate(aphia = AphiaID) %>% 
  group_by(aphia) %>% 
  group_modify(~ get_species_habitats(sp_id = .$AphiaID)) %>% 
  ungroup() %>% 
  dplyr::select(-aphia)

```
To neaten up this output, gather together similar columns:
```{r}
sp_habitat_summaries <- sp_habitat_summaries %>%
  dplyr::select(AphiaID:mean_ab, inf_epi,
                TotalD50, log_D50, GravelD50, GravelPercent,
                MudPercent:SandPercent,
                starts_with("Energy"),
                starts_with("Biozone"),
                starts_with("Substrate"),
                starts_with("Salinity"),
                starts_with("EUNIS"),
                starts_with("MSFD"),
                everything())
```

For comparative purposes, it is also useful to have summaries of the distributions and frequencies of habitat types across all sampling events. This loads a function to do that:
```{r}
source(here::here("scripts", "get_event_habitats.R"))
```

So means / frequencies of different habitat types are:
```{r}
event_habitat_summaries <- get_event_habitats()
```

We can write these two data products to file:

```{r}
write_csv(sp_habitat_summaries,
          here::here(
            "product", "benthic_specicies_habitat_summaries.csv"))
write_csv(event_habitat_summaries,
          here::here("product", "benthic_sampling_event_habitat_summaries.csv"))
```

There are a lot of variables in the summary datasets. This provides some meta-data - a three-column dataset with the variable name, a brief description, and its source. For additional information on the EUNIS classifications, see https://www.eea.europa.eu/data-and-maps/data/eunis-habitat-classification.

```{r}
sp_habitat_summaries_meta <- tibble(
  variable_name = names(sp_habitat_summaries),
  variable_description = c(
    "Taxon AphiaID",
    "Total number of sampling events the species occurred in",
    "Total abundance of the species across all sampling events",
    "Mean abundacne of the species across the sampling events in which it occurred (i.e. excluding zeros)",
    "Information on species habitat position, Infaunal, Epifaunal, or Both for species in Biotic, NA otherwise",
    "Mean of the median grain size of the whole sediment across all occurrences matched to sediment data, weighted by abundance (mm)",
    "Mean of the logged (base 10) median grain size of the whole sediment across all occurrences matched to sediment data, weighted by abundance (mm)",
    "Mean of the median grain size of the gravel fraction of sediment across all occurrences matched to sediment data, weighted by abundance (mm)",
    "Mean percentage of surface sediment on seabed composed of gravel across all occurrences matched to sediment data, weighted by abundance",
    "Mean percentage of surface sediment on seabed composed of mud across all occurrences matched to sediment data, weighted by abundance",
    "Mean percentage of area made up of surface rock across all occurrences matched to sediment data, weighted by abundance",
    "Mean percentage of area made up of rock in the top 50cm across all occurrences matched to sediment data, weighted by abundance",
    "Mean of the median grain size of the sand fraction of sediment across all occurrences matched to sediment data, weighted by abundance (mm)",
    "Mean percentage of surface sediment on seabed composed of sand across all occurrences matched to sediment data, weighted by abundance",
    "Abundance-weighted frequency of occurrence in moderate energy habitats",
    "Abundance-weighted frequency of occurrence in high energy habitats",
    "Abundance-weighted frequency of occurrence in low energy habitats",
    "Abundance-weighted frequency of occurrence in habitats with missing energy classification",
    "Abundance-weighted frequency of occurrence in habitats with no energy classification",
    "Abundance-weighted frequency of occurrence in the deep circalittoral biozone",
    "Abundance-weighted frequency of occurrence in the infralittoral biozone",
    "Abundance-weighted frequency of occurrence in the shallow circalittoral biozone",
    "Abundance-weighted frequency of occurrence in the habitats with missing biozone information",
    "Abundance-weighted frequency of occurrence in the Arctic lower bathyal biozone",
    "Abundance-weighted frequency of occurrence in the Arctic mid bathyal biozone",
    "Abundance-weighted frequency of occurrence in the Atlantic upper bathyal biozone",
    "Abundance-weighted frequency of occurrence in the Atlanto-Arctic upper bathyal biozone",
    "Abundance-weighted frequency of occurrence in the Arctic upper abyssal biozone",
    "Abundance-weighted frequency of occurrence in coarse substrate",
    "Abundance-weighted frequency of occurrence in mixed substrate",
    "Abundance-weighted frequency of occurrence in sand substrate",
    "Abundance-weighted frequency of occurrence in habitats with missing substrate information",
    "Abundance-weighted frequency of occurrence in fine mud substrate",
    "Abundance-weighted frequency of occurrence in sandy mud substrate",
    "Abundance-weighted frequency of occurrence in fine mud or muddy sand substrate",
    "Abundance-weighted frequency of occurrence in muddy sand substrate",
    "Abundance-weighted frequency of occurrence on rock or other hard substrate",
    "Abundance-weighted frequency of occurrence in sandy mud or muddy sand substrate",
    "Abundance-weighted frequency of occurrence on seabed substrate",
    "Abundance-weighted frequency of occurrence in sediment substrate",
    "Abundance-weighted frequency of occurrence in habitats with missing salinity information",
    "Abundance-weighted frequency of occurrence in mesohaline habitats",
    "Abundance-weighted frequency of occurrence in oligohaline habitats",
    "Abundance-weighted frequency of occurrence in polyhaline habitats",
    "Abundance-weighted frequency of occurrence in euhaline habitats",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.15",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.13",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.14",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.27",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.45",
    "Abundance-weighted frequency of occurrence in habitats with missing EUNIS 2007 codes",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A6.5",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A3.5",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A3.6",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.6",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.23",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.23 or A5.24",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.24",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.24 or A5.33 or A5.34",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.25",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.26 or A5.35 or A5.36",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.33",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.34",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.36",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.37",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.43",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.44",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.1",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.35",
    "Abundance-weighted frequency of occurrence in habitats with missing EUNIS 2007 codes",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.2",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.25 or A5.26",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A6",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A6.2",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A6.3 or A6.4",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A3.2",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A3.1",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A6.11",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A3.3",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.3",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.27",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.26",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.33",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.5",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4.4",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A3.4",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A5.27 or A5.37",
    "Abundance-weighted frequency of occurrence in EUNIS 2007 habitat A4",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type offshore circalittoral coarse sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type circalittoral coarse sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type infralittoral coarse sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type offshore circalittoral mixed sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type offshore circalittoral sand",
    "Abundance-weighted frequency of occurrence in habitats with missing MSFD Benthic Broad Habitat Type information",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type lower bathyal sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type circalittoral mixed sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type circalittoral mud",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type circalittoral mud or circalittoral sand",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type circalittoral rock and biogenic reef",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type circalittoral sand",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type infralittoral mixed sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type infralittoral mud",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type infralittoral mud or infralittoral sand",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type infraalittoral rock and biogenic reef",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type infraalittoral sand",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type offshore circalittoral mud",
    "Abundance-weighted frequency of occurrence in habitats with missing MSFD Benthic Broad Habitat Type information",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type upper bathyal sediment",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type upper bathyal rock and biogenic reef",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type offshore circalittoral rock and biogenic reef",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type abyssal",
    "Abundance-weighted frequency of occurrence in MSFD Benthic Broad Habitat Type offshore circalittoral mud or offshore circalittoral sand",
    "Abundance-weighted frequency of occurrence in habitats with missing oxygen concentration information",
    "Does the species have planktonic larvae? Yes / No / Unknown for species in Biotic, NA otherwise",
    "Does the species occur on bedrock? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on cobbles? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on large or very large boulders? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on small boulders? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on mud? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on muddy gravel? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on muddy sand? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on sandy mud? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on coarse sand? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on fine sand? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on other species? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on pebbles? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on artificial substrates? 1 if yes in Biotic, NA otherwise",
    "Does the species occur in caves? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on overhangs? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on gravel shingle? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on algae? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on mixed sediments? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on salt marsh? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on seagrass? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on biogenic reefs? 1 if yes in Biotic, NA otherwise",
    "Does the species occur under boulders? 1 if yes in Biotic, NA otherwise",
    "Does the species occur in crevices? 1 if yes in Biotic, NA otherwise",
    "Does the species occur in rockpools? 1 if yes in Biotic, NA otherwise",
    "Does the species occur in the pelagic zone? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on muddy gravelly sand? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on sandy gravelly mud? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on maerl beds? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on clay? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on gravelly sand? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on muddy sandy gravel? 1 if yes in Biotic, NA otherwise",
    "Does the species occur on the strandline? 1 if yes in Biotic, NA otherwise"
  ),
  source = c("WoRMS", rep("Derived", 3), "Biotic", rep("Wilson et al.", 9),
             rep("EMODnet broadscale habitat map", 100), rep("Biotic", 33))
)

```

Write this to file:
```{r}
write_csv(sp_habitat_summaries_meta,
          here::here(
            "product", "benthic_specicies_habitat_summaries_metadata.csv"))

```

The final function here produces a series of summary plots for a given species:
```{r}
source(here::here("scripts", "plot_species_habitats.R"))
```
This requires a species Aphia ID (`sp_id`). Other arguments have sensible defaults. You can print the compiled plot to your device (set `print_plot = TRUE`) - though beware, this is unlikely to look good unless your graphics device window is large. You can also save the plot to file (set `save_plot = TRUE`) - this will create a subdirectory within the 'product' directory called species_hab_plots (if it does not already exist), and save a species plot as an A4 pdf file, with the filename starting with the species Aphia ID and ending with 'habitat_plot'. Try for one species:
```{r}
plot_species_habitats(sp_id = 103228, print_plot = TRUE, save_plot = TRUE)
```
  
To create and save these composite plots for all species:

```{r, eval = FALSE}
invisible(
  sp_habitat_summaries %>%
    filter(total_occ > 20) %>% 
    mutate(aphia = AphiaID) %>% 
    group_by(aphia) %>%
    group_map(~ plot_species_habitats(sp_id = .$AphiaID,
                                      print_plot = FALSE,
                                      save_plot = TRUE,
                                      replace_plot = FALSE))
)

```

## Reproducibility
<details><summary>Reproducibility receipt</summary>
```{r}
## datetime
Sys.time()
## repository
git2r::repository()
## session info
sessionInfo()
```

