---
title: "species and traits"
author: "Tom Webb"
date: "17/06/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A dataset of substrate preferences for European benthic invertebrates

This document describes the steps taken to create a single dataset of substrate preferences for a selection of European benthic invertebrates, using data from MarLIN's [BIOTIC](http://www.marlin.ac.uk/biotic/) database, supplemented with taxonomic and species attribute data from [WoRMS](http://www.marinespecies.org) and occurrence data from [OBIS](https://obis.org).

First step is to load required packages
```{r message = FALSE}
library(tidyverse)
library(worrms)
library(robis)
library(here)
```

Now read in datasets - these are datasets including all information on individual traits, supplied from BIOTIC by Dan Lear in June 2020:
```{r, message = FALSE}
env_pos <- read_csv(here::here("data", "raw_data/biotic/bioticEnvPos.csv"))
feed_method <- read_csv(here::here("data", "raw_data/biotic/bioticFeedingMethod.csv"))
habit <- read_csv(here::here("data", "raw_data/biotic/bioticHabit.csv"))
substratum <- read_csv(here::here("data", "raw_data/biotic/bioticSubstratum.csv"))
dev_mech <- read_csv(here::here("data", "raw_data/biotic/bioticDevMech.csv"))
```

The main trait of interest is `substratum`, which includes species-level substrate preferences. To faciliate working with this, we create a lookup table with full substrate names and abbreviated names:
```{r}
substrate_values <- tibble(
  substratum = unique(substratum$substratum)
  ) %>%
  mutate(substratum_code = case_when(
    substratum == "Large to very large boulders" ~ "Large_VLarge_Boulders",
    substratum == "Small boulders" ~ "Small_Boulders",
    substratum == "Muddy gravel" ~ "Muddy_gravel",
    substratum == "Muddy sand" ~ "Muddy_sand",
    substratum == "Sandy mud" ~ "Sandy_mud",
    substratum == "Coarse clean sand" ~ "Coarse_sand",
    substratum == "Fine clean sand" ~ "Fine_sand",
    substratum == "Other species (see additional information)" ~ "Other_species",
    substratum == "Artificial (e.g. metal/wood/concrete)" ~ "Artificial",
    substratum == "Insufficient information" ~ "No_Info",
    substratum == "Gravel / shingle" ~ "Gravel_shingle",
    substratum == "Salt marsh" ~ "Salt_marsh",
    substratum == "Biogenic reef" ~ "Biogenic_reef",
    substratum == "Under boulders" ~ "Under_boulders",
    substratum == "Crevices / fissures" ~ "Crevices",
    substratum == "Water column (pelagic)" ~ "Pelagic",
    substratum == "No preference" ~ "No_preference",
    substratum == "Muddy gravelly sand" ~ "Mud_grav_sand",
    substratum == "Sandy gravelly mud" ~ "Sand_grav_mud",
    substratum == "See additional information" ~ "No_Info",
    substratum == "Gravelley sand" ~ "Gravel_sand",
    substratum == "Muddy sandy gravel" ~ "Mud_sand_gravel",
    TRUE ~ substratum
  ))
```
Then replace long values with abbreviated values in substratum dataframe:
```{r}
substratum <- left_join(substratum, substrate_values, by = "substratum") %>% 
  select(-substratum)
```
We can remove uninformative records (no info or no preference):
```{r}
substratum <- substratum %>% filter(!(substratum_code %in% c("No_Info", "No_preference")))
```
Now create a wide version of this data (one species per row, each substrate forms a column) - do this for each trait:
```{r}
substratum_wide <- substratum %>% 
  count(SpeciesName, substratum_code) %>%
  pivot_wider(names_from = substratum_code, values_from = n)

env_pos_wide <- env_pos %>%
  count(SpeciesName, envpos) %>%
  pivot_wider(names_from = envpos, values_from = n)

feed_method_wide <- feed_method %>% 
  count(SpeciesName, feedingmethod) %>%
  pivot_wider(names_from = feedingmethod, values_from = n)

habit_wide <- habit %>% 
  count(SpeciesName, Habit) %>%
  pivot_wider(names_from = Habit, values_from = n)

```

Use these to get a full species list:
```{r}
biotic_species <- unique(c(
  pull(env_pos_wide, SpeciesName),
  pull(feed_method_wide, SpeciesName),
  pull(habit_wide, SpeciesName),
  pull(substratum_wide, SpeciesName)
))
```
To match the species names to the WoRMS taxonomy, we could use the `worrms` package:
```{r, eval = FALSE}
biotic_aphias <- wm_name2id_(biotic_species)
biotic_aphias <- biotic_aphias %>% enframe() %>% mutate(value = unlist(value))
```
However, a significant proportion do not result in a match, so instead we use the WoRMS taxon match tool at http://www.marinespecies.org/aphia.php?p=match - first writing the species list to file, 
```{r, eval = FALSE}
write_csv(tibble(name = biotic_species), here::here("data", "derived_data/biotic_aphias_to_check.txt"))
```
Then reading in the matched records:
```{r, message = FALSE, warning = FALSE}
biotic_aphias_matched <- read_tsv(here::here("data", "derived_data/biotic_aphias_to_check_matched.txt"))
```
Next remove unmatched species, plus any at rank above species
```{r}
biotic_aphias_matched <- biotic_aphias_matched %>% 
  filter(!is.na(AphiaID_accepted) & !is.na(Species)) %>% 
  select(-c(Subgenus, Species, Subspecies))
```
Now add functional group, using functional group matching function in `get_worms_fgrp.R` - load this function first (also available from https://github.com/tomjwebb/WoRMS-functional-groups):
```{r}
source(here::here("scripts", "get_worms_fgrp.R"))
```
Then run it:
```{r}
spp_attr <- biotic_aphias_matched %>%
  mutate(aphia = AphiaID_accepted) %>% 
  group_by(AphiaID_accepted) %>% 
  group_map(~ get_worms_fgrp(AphiaID = .x$aphia)) %>% 
  bind_rows()
```
Do a quick check of species with two adult classifications:
```{r}
spp_attr %>% filter(!is.na(adult_2))
```
And any that have two larval classifications:
```{r}
spp_attr %>% filter(!is.na(`larva > planula_2`))
```
Or both larva and `larva > planula`:
```{r}
spp_attr %>% filter(!is.na(larva) & !is.na(`larva > planula`))
```
Or both polyp and adult:
```{r}
spp_attr %>% filter(!is.na(polyp) & !is.na(adult))
```
How many species have juvenile data?
```{r}
spp_attr %>% filter(!is.na(juvenile))
```
Given this we can create a simplified version of just adult and larval functional group:
```{r}
spp_attr_simple <- spp_attr %>%
  mutate(larva = ifelse(is.na(larva), `larva > planula`, larva),
         adult = ifelse(is.na(adult), polyp, adult)) %>% 
  select(AphiaID, adult, larva)
```
Next step, get an OBIS checklist for each species - this will show how many (global) occurrence records there are for species:
```{r, message = FALSE}
spp_obis <- checklist(taxonid = biotic_aphias_matched$AphiaID_accepted) %>% 
  as_tibble()
```
Note - this returns AphiaIDs that are not in our biotic Aphias dataset:
```{r}
spp_obis %>% filter(!(taxonID %in% biotic_aphias_matched$AphiaID_accepted))
```
All of these are at ranks below species level:
```{r}
spp_obis %>% filter(!(taxonID %in% biotic_aphias_matched$AphiaID_accepted)) %>% count(taxonRank)
```
They also include rather few total OBIS records, and can just be removed here in the next step, which simplifies to just AphiaID and OBIS records:
```{r}
spp_obis_simple <- spp_obis %>%
  filter(taxonID %in% biotic_aphias_matched$AphiaID_accepted) %>% 
  select(taxonID, records) %>% 
  rename(AphiaID = taxonID, obis_records = records)
```
Now join the OBIS and the Functional Group data:
```{r}
spp_fg_obis <- full_join(spp_attr_simple, spp_obis_simple, by = "AphiaID")
```
Now add taxonomic info back into substratum data, then add the Functional Group and OBIS info:
```{r}
substratum_full <- substratum_wide %>% left_join(
  select(biotic_aphias_matched, name, AphiaID, AphiaID_accepted, ScientificName_accepted, Kingdom:Family),
  by = c("SpeciesName" = "name")) %>% 
  filter(Kingdom == "Animalia") %>% 
  select(-c(Kingdom)) %>% 
  left_join(spp_fg_obis, by = c("AphiaID_accepted" = "AphiaID")) %>% 
  select(AphiaID_accepted, ScientificName_accepted, adult, larva, obis_records, everything())
```
Restrict to benthos (broad sense):
```{r}
substratum_full <- substratum_full %>%
  filter(adult %in% c("benthos", "endobenthos", "epibenthos", "macrobenthos"))
```
Add in other traits - first Environmental Position (infanua / epifauna). Create a simple version of `env_pos` first  - there are quite a few classifications here:
```{r}
env_pos %>% count(envpos)
```
But most can be simplified to inf/epi. So:
```{r}
env_pos_simple <- env_pos %>% 
  left_join(select(biotic_aphias_matched, name, AphiaID_accepted), by = c("SpeciesName" = "name")) %>% 
  filter(AphiaID_accepted %in% substratum_full$AphiaID_accepted) %>% 
  mutate(envpos_simple = ifelse(str_detect(envpos, "Epi"), "Epifaunal", envpos)) %>% 
  count(AphiaID_accepted, envpos_simple) %>%
  mutate(n = 1) %>% 
  pivot_wider(names_from = envpos_simple, values_from = n) %>% 
  rowwise() %>% 
  mutate(n_pos = sum(Epifaunal, Infaunal, Demersal, Interstitial, Lithotomous, na.rm = TRUE)) %>% 
  select(AphiaID_accepted, n_pos, everything()) %>% 
  ungroup()
```
Check species with multiple classifications:
```{r}
env_pos_simple %>% filter(n_pos > 1)
```
For all of these we can use infaunal / epifaunal or 'both'. Quick check of species with only one value, in one of Demersal, Interstitial, Lithotomous:
```{r}
env_pos_simple %>% filter(n_pos == 1 & (Demersal == 1 | Interstitial == 1 | Lithotomous == 1))
```
107387 is _Liocarcinus depurator_, swimming crab, listed in Biotic as swimmer/crawler/burrower so inf/epi seems appropriate. 138945 is _Caecum armoricum_, a gastropod, interstitial which we will class as infauna. 140770 is _Pholas dactylus_, a burrowing mollusc (common piddock), which burrows into soft rock / artificial structures. Hard substrate (bedrock) so classing here as epifauna. Update these, and simplify to infauna/epifauna/both:
```{r}
env_pos_simple <- env_pos_simple %>%
  mutate(Epifaunal = ifelse(AphiaID_accepted %in% c(107387, 138945), 1, Epifaunal),
                          Infaunal = ifelse(AphiaID_accepted %in% c(107387, 140770), 1, Infaunal),
                          inf_epi = case_when(
                            Epifaunal == 1 & Infaunal == 1 ~ "Both",
                            Epifaunal == 1 & is.na(Infaunal) ~ "Epifaunal",
                            Infaunal == 1 & is.na(Epifaunal) ~ "Infaunal"
                          )
                          ) %>% 
  select(AphiaID_accepted, inf_epi)
```
Add to substratum data:
```{r}
substratum_full <- substratum_full %>% 
  left_join(env_pos_simple, by = "AphiaID_accepted") %>% 
  select(AphiaID_accepted:obis_records, inf_epi, everything())
```
Create a simple version of dev_mech (larval development mode). There are quite a few classifications here
```{r}
dev_mech %>% count(devmech)
```
Restrict to species in the substratum dataset, and simplify dev mech where possible. NB - viviparity / oviparity tells us nothing about larval dispersal (e.g. _Bugula turbinata_ is listed as viviparous, but also has lecithotrophic larvae - see http://www.marlin.ac.uk/biotic/browse.php?sp=4418).
```{r}
dev_mech_simple <- dev_mech %>% 
  left_join(select(biotic_aphias_matched, name, AphiaID_accepted), by = c("SpeciesName" = "name")) %>% 
  filter(AphiaID_accepted %in% substratum_full$AphiaID_accepted) %>% 
  mutate(devmech_simple = case_when(
    devmech %in% c("Brooding", "Direct Development", "Schizotomous") ~ "Direct",
    devmech %in% c("Lecithotrophic", "Planktotrophic") ~ "Planktonic",
    TRUE ~ "Other"
  )) %>% 
  count(AphiaID_accepted, devmech_simple) %>%
  mutate(n = 1) %>% 
  pivot_wider(names_from = devmech_simple, values_from = n) %>% 
  rowwise() %>% 
  mutate(n_devmech = sum(Planktonic, Direct, Other, na.rm = TRUE)) %>% 
  select(AphiaID_accepted, n_devmech, everything()) %>% 
  ungroup()
```
Check species with planktonic AND direct classifications:
```{r}
dev_mech_simple %>% filter(Planktonic == 1 & Direct == 1)
```
This includes things like _Cordylophora caspia_ which have both sexual (planktonic) and asexual (direct) reproduction. For our purposes - what's important is if larvae *might* turn up in a plankton survey, so we want a planktonic yes/no/unknown variable only, and can just join that to substratum data.
```{r}
dev_mech_simple <- dev_mech_simple %>%
  mutate(plank_biotic = case_when(
  Planktonic == 1 ~ "Yes",
  Direct == 1 & is.na(Planktonic) ~ "No",
  TRUE ~ "Unknown"))
```
Add to substratum data:
```{r}
substratum_full <- substratum_full %>% 
  left_join(select(dev_mech_simple, AphiaID_accepted, plank_biotic), by = "AphiaID_accepted") %>% 
  mutate(plank_larv = case_when(
    larva == "zooplankton" | plank_biotic == "Yes" ~ "Yes",
    plank_biotic == "No" ~ "No",
    TRUE ~ "Unknown")) %>% 
  select(AphiaID_accepted, ScientificName_accepted, obis_records, inf_epi, plank_larv,
         Bedrock:Strandline, everything())
```
Some final tidying to remove/rename columns:
```{r}
substratum_full <- substratum_full %>% select(-c(adult, larva, plank_biotic)) %>% 
  rename(biotic_SpeciesName = SpeciesName, biotic_AphiaID = AphiaID)
```
Now export this derived dataset, together with the substrate values table:
```{r}
write_csv(substratum_full, here::here("data", "derived_data/benthic_species_substratum_prefs.csv"))
write_csv(substrate_values, here::here("data", "derived_data/substrate_values_key.csv"))
```
